---
title: "KM_Plot_In_Class_Work"
author: Jeanne Revilla, adapted from Nicole Black and Kayla Xu
date: "2024-09-29"
output: pdf_document
---

### (0) Setting Up!

##### Setting Your Working Directory
```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("~/490_cluster/analysis_data")) 
```

##### Download and Load Packages / Data
```{r}

if (!require("BiocManager")) install.packages("BiocManager")

if (!require("TCGAbiolinks")) BiocManager::install("TCGAbiolinks")

if (!require("survival")) install.packages("survival")

if(!require("survminer")) install.packages("survminer")

if(!require("ggplot2")) install.packages("ggplot2")

library(BiocManager) 
library(TCGAbiolinks)
library(survival)
library(survminer)
library(ggplot2)
```

If you don't already have the data, run this code!
```{r}
clin_query <- GDCquery(project = "TCGA-BRCA", data.category = "Clinical",
data.type = "Clinical Supplement", data.format = 'BCR Biotab')

#GDCdownload(clin_query)

clinical.BCRtab.all <- GDCprepare(clin_query)
clinical <- clinical.BCRtab.all$clinical_patient_brca[-c(1,2),]
```


### (1) Cleaning and Prepping the Data
For today, we will be creating a KM Plot by Age Category for BRCA data.

##### Clean the age_at_diagnosis column
```{r}
# The relevant column is age_at_diagnosis

age_NA_mask <- !is.na(clinical$age_at_diagnosis)

clean_clinical <- clinical[age_NA_mask,]

```

##### Create the age_category column. 
```{r}
clean_clinical$age_category <- ifelse(clean_clinical$age_at_diagnosis < 40, "Young", ifelse(clean_clinical$age_at_diagnosis >=65, "Old", "Middle"))

```

##### Create the survival_time column.
```{r}
# There are two important columns for this step:
#         1. days_to_last_followup
#         2. death_days_to

clean_clinical$survival_time <- ifelse(clean_clinical$death_days_to == "[Not Applicable]", clean_clinical$last_contact_days_to, clean_clinical$death_days_to)


clean_clinical$survival_time <- as.numeric(clean_clinical$survival_time)

```

##### Make the death_event column.
```{r}
#The relevant column  is vital_status

clean_clinical$death_event <- ifelse(clean_clinical$vital_status == "Dead", T, F)

```


### (2) Create Survminer Objects
```{r}
survival_object <- Surv(time = clean_clinical$survival_time, 
                        event = clean_clinical$death_event)

fit_object <- survfit(survival_object ~ clean_clinical$age_category, 
                      data = clean_clinical)
  
```


### (3) Create the Plot
```{r}
survplot <- ggsurvplot(fit_object, pval = TRUE,
                       ggtheme = theme(plot.margin 
                                       = unit(c(1,1,1,1), "cm")), 
                       legend = "right")

KM_plot <- survplot$plot + theme_bw() + 
  theme(axis.title = element_text(size=20),
        axis.text = element_text(size=16),
        legend.title = element_text(size=14),
        legend.text = element_text(size=12))

```


### (4) View your Plot!
```{r}
KM_plot

```

